---
title: "Artificial Bee Colony Optimization"
author: "Jose Caloca"
date: "14/5/2021"
output: html_document
---

# Artificial Bee Colony Optimization:

Implements Karaboga (2005) Artificial Bee Colony (ABC) Optimization algorithm.

## Parameters

param par Initial values for the parameters to be optimized over

param fn A function to be minimized, with first argument of the vector of parameters over which minimization is to take place. It should return a scalar result.

param ... In the case of \code{abc_*}, further arguments to be passed to 'fn', otherwise, further arguments passed to the method.

param FoodNumber Number of food sources to exploit. Notice that the

param \code{NP} has been deprecated.

param lb Lower bound of the parameters to be optimized.

param ub Upper bound of the parameters to be optimized.

param limit Limit of a food source.

param maxCycle Maximum number of iterations.

param optiinteger Whether to optimize binary parameters or not.

param criter Stop criteria (numer of unchanged results) until stopping param parscale Numeric vector of length **length(par)**. Scale applied to the parameters.

param fnscale Numeric scalar. Scale applied function. If **fnscale \< 0**, then the problem becomes a maximization problem.

## details

Upper and Lower bounds (***ub, lb***) equal to infinite will be replaced by either ***.Machine\$double.xmax or -.Machine\$double.xmax***

If ***D*** (the number of parameters to be optimzed) is greater than one, then ***lb*** and ***ub*** can be either scalars (assuming that all the parameters share the same boundaries) or vectors (the parameters have different boundaries each other).

return An list of class ***abc_answer***, holding the following elements:

Foods : {Numeric matrix. Last position of the bees.}

f: {Numeric vector. Value of the function evaluated at each set of Foods.}

fitness: {Numeric vector. Fitness of each Foods.}

trial :{Integer vector. Number of trials at each \code{Foods}.}

value :{Numeric scalar. Value of the function evaluated at the optimum.}

par: {Numeric vector. Optimum found.}

counts: {Integer scalar. Number of cycles.}

hist: {Numeric matrix. Trace of the global optimums.}

# Algorithm definition

```{r}
artificial_bee_colony <- function(
    par,               # Vector with parameters to optimise 
    fn,                # Objective function
    ...,               # Arguments of the function (M, x0, X, etc.)
    FoodNumber  = 20,   # Fuentes de alimento 
    lb          = rep(-Inf, length(par)),        # Lower limit of the distance
    ub          = rep(+Inf, length(par)),        # Upper limit of the distance
    limit       = 100,       # Limit how depleted a food source is
    maxCycle    = 1000,   # Maximum number of iterations 
    optiinteger = FALSE, # TRUE if we want to optimize on [0,1] (binary)
    criter      = 50,
    parscale    = rep(1, length(par)),
    fnscale     = 1
)
{
    D <- length(par)
    
    # Checking limits
    if (length(lb) == 1 && length(par) > 1) lb <- rep(lb, D)
    if (length(ub) == 1 && length(par) > 1) ub <- rep(ub, D)
    
    lb[is.infinite(lb)] <- -.Machine$double.xmax*1e-10
    ub[is.infinite(ub)] <- .Machine$double.xmax*1e-10
    
    # Initial params
    Foods       <- matrix(double(FoodNumber*D), nrow=FoodNumber)
    f           <- double(FoodNumber)
    fitness     <- double(FoodNumber)
    trial       <- double(FoodNumber)
    prob        <- double(FoodNumber)
    solution    <- double(D)
    ObjValSol   <- double(1)
    FitnessSol  <- double(1)
    neighbour   <- integer(1)
    param2change<- integer(1)
    GlobalMin   <- fn(par, ...) # double(1)
    GlobalParams<- par #double(D)
    #GlobalMins  <- double(runtime)
    r           <- integer(1)
    
    # Fun
    fun <- function(par) fn(par/parscale, ...)/fnscale
    
    # Fitness function
    CalculateFitness <- function(fun)
    {
        if (fun >= 0) return(1/(fun + 1))
        else return(1 + abs(fun))
    }
    # CalculateFitness(f[1])
    
    # The best food source is memorized
    MemorizeBestSource <- function() 
    {
        oldGlobalMin <- GlobalMin
        for(i in seq(1,FoodNumber)) {
            if (f[i] < GlobalMin) {
                GlobalMin <<- f[i]
                
                # Replacing new group of parameters
                GlobalParams <<- Foods[i,]
            }
        }
        
        # Increasing persistance
        if (oldGlobalMin == GlobalMin) persistance <<- persistance + 1
        else persistance <<- 0
    }
    
    # Variables are initialized in the range [lb,ub]. If each parameter has
    # different range, use arrays lb[j], ub[j] instead of lb and ub 
    # Counters of food sources are also initialized in this function
    
    init <- function(index, ...) {
        if (optiinteger) Foods[index,] <<- runif(D) > .5
        else {
            Foods[index,] <<- sapply(1:D, function(k) runif(1,lb[k],ub[k]) )
        }
        
        solution <<- Foods[index,]
        
        f[index] <<- fun(solution)
        
        fitness[index] <<- CalculateFitness(f[index])
        trial[index] <<- 0
        
    }
    # init(2)
    
    # All food sources are initialized
    initial <- function() {
        # For the first initialization we set the bees at
        # specific places equaly distributed through the
        # bounds.
        Foods <<- 
            sapply(1:D, function(k) {
                seq(lb[k],ub[k],length.out=FoodNumber)
            }
            )
        
        for (i in 1:FoodNumber) {
            solution <<- Foods[i,]
            
            f[i] <<- fun(solution)
            
            fitness[i] <<- CalculateFitness(f[i])
            trial[i] <<- 0
        }
    }
    
    # initial()
    
    
    SendEmployedBees <- function() {
        for (i in 1:FoodNumber) {
            # The parameter to be changed is determined randomly
            param2change <- sample(1:D, 1) # floor(runif(1)*D) + 1 
            
            # A randomly chosen solution is used in producing a mutant solution of the solution i
            # Randomly selected solution must be different from the solution i
            neighbour <- i
            while(neighbour==i)
                neighbour <- sample(1:FoodNumber, 1) # floor(runif(1)*FoodNumber) + 1
            
            solution <<- Foods[i,]
            
            # v_{ij}=x_{ij}+\phi_{ij}*(x_{kj}-x_{ij}) 
            
            if (optiinteger) solution[param2change] <<- runif(1) > 0.5
            else {
                solution[param2change] <<- 
                    Foods[i,param2change]+
                    (Foods[i,param2change]-Foods[neighbour,param2change])*(runif(1)-0.5)*2
                
                # if generated parameter value is out of boundaries, it is shifted onto the boundaries
                if (solution[param2change]<lb[param2change])
                    solution[param2change]<<-lb[param2change]
                
                if (solution[param2change]>ub[param2change])
                    solution[param2change]<<-ub[param2change]
            }
            
            ObjValSol <<- fun(solution)
            FitnessSol <<- CalculateFitness(ObjValSol)
            
            # a greedy selection is applied between the current solution i and its mutant*/
            if (FitnessSol>fitness[i]) {
                # If the mutant solution is better than the current solution i, replace the solution with the mutant and reset the trial counter of solution i*/
                trial[i] <<- 0;
                #for(j in 1:D) Foods[i,j] <<- solution[j]
                Foods[i,] <<- solution
                f[i]<<- ObjValSol
                fitness[i]<<-FitnessSol
            }
            else {
                # the solution i can not be improved, increase its trial counter*/
                trial[i] <<- trial[i]+1
            }
        }
    }
    
    
    # A food source is chosen with the probability which is proportioal to its quality*/
    # Different schemes can be used to calculate the probability values*/
    # For example prob(i)=fitness(i)/sum(fitness)*/
    # or in a way used in the metot below prob(i)=a*fitness(i)/max(fitness)+b*/
    # probability values are calculated by using fitness values and normalized by dividing maximum fitness value*/
    CalculateProbabilities <- function() {
        maxfit <- fitness[1]
        for (i in 1:FoodNumber) 
            if (fitness[i] > maxfit) maxfit <- fitness[i]
        
        prob <<- .9*(fitness/(maxfit+1e-20)) + .1
        #     prob[is.nan(prob)]  <<- .1
    }
    
    SendOnlookerBees <- function()
    {
        # Onlooker Bee phase
        i <- 1
        t <- 0
        while (t < FoodNumber)
        {
            
            # choose a food source depending on its probability to be chosen
            if (runif(1) < prob[i]) {
                t <- t + 1
                
                # The parameter to be changed is determined randomly
                param2change <- sample(1:D, 1) # floor(runif(1)*D) + 1 
                
                # A randomly chosen solution is used in producing a mutant solution of the solution i
                #Randomly selected solution must be different from the solution i*/        
                neighbour <- i
                while(neighbour==i)
                    neighbour <- sample(1:FoodNumber, 1) # floor(runif(1)*FoodNumber) + 1
                
                solution <<- Foods[i,]
                
                # v_{ij}=x_{ij}+\phi_{ij}*(x_{kj}-x_{ij}) */
                
                if (optiinteger) solution[param2change] <<- runif(1) > .5
                else 
                {
                    solution[param2change] <<- 
                        Foods[i,param2change]+
                        (Foods[i,param2change]-Foods[neighbour,param2change])*(runif(1)-0.5)*2
                    
                    # if generated parameter value is out of boundaries, it is shifted onto the boundaries*/
                    if (solution[param2change]<lb[param2change]) 
                        solution[param2change] <<- lb[param2change]
                    
                    if (solution[param2change]>ub[param2change]) 
                        solution[param2change] <<- ub[param2change]
                    
                }
                
                ObjValSol <<- fun(solution)
                FitnessSol <<- CalculateFitness(ObjValSol)
                
                # a greedy selection is applied between the current solution i and its mutant*/
                if (FitnessSol>fitness[i])
                {
                    # If the mutant solution is better than the current solution i, replace the solution with the mutant and reset the trial counter of solution i*/
                    trial[i] <<- 0
                    Foods[i,] <<- solution
                    
                    f[i]<<-ObjValSol
                    fitness[i]<<-FitnessSol
                } #if the solution i can not be improved, increase its trial counter*/
                else trial[i] <<- trial[i]+1
            }
            i <- i + 1
            if (i==FoodNumber) i <- 1
            # end of onlooker bee phase
        }
    }
    
    # determine the food sources whose trial counter exceeds the "limit" value.
    # In Basic ABC, only one scout is allowed to occur in each cycle*/
    
    SendScoutBees <- function() {
        maxtrialindex <- 1
        for (i in 1:FoodNumber) {
            if (trial[i] > trial[maxtrialindex]) maxtrialindex <- i
        }
        
        if (trial[maxtrialindex] >= limit) init(maxtrialindex)
    }
    
    persistance <- 0
    
    # Inicializa funcion
    initial()
    
    # Memoriza la primera mejor solucion
    MemorizeBestSource() 
    
    ans  <- matrix(0, ncol = D, nrow=maxCycle)
    iter <- 0
    # Comienza a iterar
    while ((iter <- iter + 1) < maxCycle)
    {
        SendEmployedBees()
        CalculateProbabilities()
        SendOnlookerBees() 
        MemorizeBestSource()
        
        # Storing parameter and breaking out
        ans[iter,] <- GlobalParams
        if (persistance > criter) break
        
        SendScoutBees()
    }
    
    return(
        structure(list(
            Foods   = Foods,
            f       = f,
            fn      = fn,
            fitness = fitness,
            trial   = trial,
            value   = fun(GlobalParams),
            par     = GlobalParams,
            counts  = c("function"=iter),
            hist    = ans[1:iter,,drop=FALSE]
        ), class="abc_answer"
        ))

}

```

# Algorithm implementation

## **Example 1**: The minimum is at (pi,pi)

```{r}
# Minimisation of the objective function
fun <- function(x) {
  -cos(x[1])*cos(x[2])*exp(-((x[1] - pi)^2 + (x[2] - pi)^2))
}

ans <- artificial_bee_colony(rep(0,2), fun, lb=-10, ub=10, criter=50)
ans[c("par", "counts", "value")]
```

```{r}
# Maximisation of the objective function 

fun <- function(x) {
  # We've removed the '-' from the equation
  cos(x[1])*cos(x[2])*exp(-((x[1] - pi)^2 + (x[2] - pi)^2))
}
ans <- artificial_bee_colony(rep(0,2), fun, lb=-10, ub=10, criter=50)
ans[c("par", "counts", "value")]

```

## **Example 2**: global minimum at about (-15.81515)

```{r}
fw <- function (x){
    10*sin(0.3*x)*sin(1.3*x^2) + 0.00001*x^4 + 0.2*x+80
}

ans <- artificial_bee_colony(50, fw, lb=-100, ub=100, criter=100)
ans[c("par", "counts", "value")]
```

## 

## **Example 3:** 5D sphere, global minimum at about (0,0,0,0,0)

```{r}
fs <- function(x) sum(x^2)

ans <- artificial_bee_colony(rep(10,5), fs, lb=-100, ub=100, criter=200)
ans[c("par", "counts", "value")]
```

## **Example 4:** An Ordinary Linear Regression

```{r}
set.seed(1231)
k <- 4
n <- 5e2

# Data generating process
w <- matrix(rnorm(k), ncol=1)     # This are the model parameters
X <- matrix(rnorm(k*n), ncol = k) # This are the controls
y <- X %*% w                      # This is the observed data

# Objective function
fun <- function(x) {
  sum((y - X%*%x)^2)
}

# Running the regression
ans <- artificial_bee_colony(rep(0,k), fun, lb = -10000, ub=10000)

# Here are the outcomes: Both columns should be the same
cbind(ans$par, w)

# This is just like OLS, with no constant
coef(lm(y~0+X))
```
